---
title: "Project Assignment: Daily Step Activity Analysis"
output: 
  html_document: 
    keep_md: TRUE
---

```{r, echo=TRUE, message=FALSE, warning=FALSE}
## SETUP ENVIRONMENT
library(dplyr)
library(sqldf)
library(lubridate)
require(data.table)
library(graphics)
library(stringr)
library(chron)
library(knitr)
library(ggplot2)
## example pre-requisite to code execution
## setwd("C:/Donald/Personal/Coursera/Data Science Specialization/Course5_Reproducible Research/Week2/Assignment")
```


```{r, echo=TRUE, message=FALSE, warning=FALSE}
## READ_RAW_FILE
activity            <- fread("./repdata-data-activity/activity.csv", sep = ",", header= TRUE)
```

******************************************************************
******************************************************************
******************************************************************
******************************************************************
******************************************************************
##Question#1: 
##What is mean total number of steps taken per day?
##Solution#1:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
activity_date 	<- sqldf('SELECT SUM(steps) AS steps , date FROM activity  WHERE steps != "NA" GROUP BY date')
hist(activity_date$steps, main="Histogram for Daily Step Count", xlab="Step Count", ylab="Frequency (Count of Days)",    border="blue")

mean_step_count 	<- as.character(round(mean(activity_date$steps), digits = 0))
median_step_count 	<- as.character(round(median(activity_date$steps), digits = 0))
```
##Conclusion#1: 
###Mean of (Total Step Count/Day): `r mean_step_count`
###Median of (Total Step Count/Day): `r median_step_count`



******************************************************************
******************************************************************
******************************************************************
******************************************************************
******************************************************************
##Question#2: 
##What is the average daily activity pattern?
##Solution#2:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
activity_int_avg    <- sqldf('SELECT AVG(steps) AS steps , interval FROM activity  WHERE steps != "NA" GROUP BY interval')
with(activity_int_avg,plot(interval, steps, main="Daily Average Step Count by Time Interval", xlab="Time of Day (HHMM)", ylab="Average Step Count", type="l", col="red"), xlim = c("0000","2359"))

max_interval 	    <- as.character(activity_int_avg$interval[which.max(activity_int_avg$steps)])
abline(v = activity_int_avg$interval[which.max(activity_int_avg$steps)], col="black", lwd=3, lty=2)
max_avg_steps	    <- as.character(max(activity_int_avg$steps))
avg_avg_steps	    <- as.character(round(mean(activity_int_avg$steps)))
```
##Conclusion#2: 
###`r max_avg_steps` represents the overall Maximum (Step Count by Interval)
###`r avg_avg_steps` represents the overall Mean (Step Count by Interval).
###`r max_interval` Interval of Day (HHMM), as indicated by the vertical dotted black line above, identifies the Interval which has the Maximum Step Count averaged on a Daily basis.

******************************************************************
******************************************************************
******************************************************************
******************************************************************
******************************************************************
##Question#3: 
##Inputing missing values
##Solution#3:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
activity_fill	            <- activity
missing_row_count_before	<- sum(is.na(activity_fill$steps))

for (i in 1:nrow(activity_fill))
	{
		if (is.na(activity_fill$steps[i]))
		{
			avg_step 	<- activity_int_avg$steps[which(activity_int_avg$interval == activity_fill$interval[i])]
			activity_fill$steps[i] 	<- avg_step
		}
	}	

missing_row_count_after		<- sum(is.na(activity_fill$steps))

activity_date_fill		    <- sqldf('SELECT SUM(steps) AS steps, date FROM activity_fill GROUP BY date')
hist(activity_date_fill$steps, main="Histogram for Daily Step Count (complete data set)", xlab="Step Count", ylab="Frequency (Count of Days)",    border="orange")

mean_step_count_fill 		<- as.character(round(mean(activity_date_fill$steps)))
median_step_count_fill 		<- as.character(median(activity_date_fill$steps))
```
##Conclusion#3: 
###Strategy to fill up NA values in the "steps" column: Average value of this variable from another matching non-NA interval was used to fill up the NA value. This gives a more accurate result.
###Mean of (Total Step Count per Day): `r mean_step_count_fill`
###Median of (Total Step Count per Day): `r median_step_count_fill`
###Mean and Median of (Total Step Count per Day) has slightly decreased after filling up missing(NA) values in the steps column. 

******************************************************************
******************************************************************
******************************************************************
******************************************************************
******************************************************************
##Question#4: 
##Are there differences in activity patterns between weekdays and weekends?
##Solution#4:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
activity_fill$isweekend     <- is.weekend(as.Date(activity_fill$date, "%Y-%m-%d"))

activity_int_avg_fill 		<- sqldf('SELECT AVG(steps) AS steps , interval, isweekend FROM activity_fill  GROUP BY interval, isweekend')

g <- qplot(interval, steps , data = activity_int_avg_fill) 
isweekend_names			    <- c(`FALSE`="weekday",`TRUE`="weekend") 	
g + labs(title="Daily Average Step Count Comparison") + labs(x="Time of Day (HHMM)", y="Avg Daily Step Count") +geom_line(color="blue")+ geom_point() + facet_grid(. ~ isweekend, labeller = as_labeller(isweekend_names))

activity_int_avg_fill_weekend <- mean(subset(activity_int_avg_fill,isweekend==TRUE)$steps)
activity_int_avg_fill_weekday <- mean(subset(activity_int_avg_fill,isweekend==FALSE)$steps)

activity_int_max_fill_weekend <- max(subset(activity_int_avg_fill,isweekend==TRUE)$steps)
activity_int_max_fill_weekday <- max(subset(activity_int_avg_fill,isweekend==FALSE)$steps)

```
##Conclusion#4:
###Average of (Step Count per Interval) on Weekday (`r activity_int_avg_fill_weekday`), is less than that  on Weekend (`r activity_int_avg_fill_weekend`).
###Max (Step Count per Interval) on Weekday (`r activity_int_max_fill_weekday`), is more than that on Weekend (`r activity_int_max_fill_weekend`).
###Activity time period starts at around 500 (HHMM) on Weekdays and Weekends.
###Activity time period ends slightly later on Weekdays, compared to Weekends.

******************************************************************
******************************************************************
******************************************************************
******************************************************************
******************************************************************